---
title: "R vs Python"
# Credit to :
# grantmcdermott/two-col-test
# For the two columns setup
author: "Claude REN"
date: "10/22/2020"
output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
    includes:
      in_header: ["latex_topmatter.tex", "header.tex"]
---

```{r setup, include=FALSE}
library(reticulate)
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In this .pdf you will find all the basics of R and its equivalence in Python. The purpose is to provide a document where you can easily find useful information when working on R and Python. 

## Basic operator

:::::: {.columns}
::: {.column width="48%" data-latex="{0.48\textwidth}"}
### Calculate with R

```{r, eval=TRUE}
# No import
10*(1+3-2.4)
10^2
10**2
sqrt(100)
pi
cos(pi)
exp(1)
log(1)
round(2.5435, 2)
a <- 100
print(a)
```
:::
::: {.column width="4%" data-latex="{0.04\textwidth}"}
\ 
<!-- an empty Div (with a white space), serving as
a column separator -->
:::
:::::: {.column width="48%" data-latex="{0.48\textwidth}"}

### Calculate with Python

```{python, eval=TRUE}
import numpy as np
10*(1+3-2.4)
10^2        # Not correct !
10**2
np.sqrt(100)   
np.pi          
np.cos(np.pi)     
np.exp(1)      
np.log(1)      
round(2.543534, 2)
a = 100
print(a)
```
:::
::::::

\newline

:::::: {.columns}
::: {.column width="48%" data-latex="{0.48\textwidth}"}

### Vector operations with R

```{r, eval=TRUE}
v <- c(10, 20, 30)
v
length(v)
2*v+1
v**2
log(v)
w <- c(1, 2, 3)
v-w
v*w
v/w
v%*%w
sum(v)
mean(v)
min(v)
max(v)
sd(v)
median(v)
```

:::
::: {.column width="4%" data-latex="{0.04\textwidth}"}
\ 
<!-- an empty Div (with a white space), serving as
a column separator -->
:::
:::::: {.column width="48%" data-latex="{0.48\textwidth}"}

### Vector operations with Python

```{python, eval=TRUE}
v = np.array([10, 20, 30])
v
len(v)
2*v+1
v**2
np.log(v).round(4)
w = np.array([1, 2, 3])
v-w
v*w
v/w
np.dot(v,w)
sum(v)
np.average(v)
min(v)
max(v)
np.std(v, ddof = 1)
np.median(v)
```

:::
::::::

\newline

For the standard deviation, the formula used in numpy with std() is different from the one in R with sd().

\begin{center}

$\sigma_p = \sqrt{\frac{1}{n}\sum_{i=1}^n (x_i-\overline{x})^2} = \sqrt{\frac{1}{n}\sum_{i=1}^n x_i^2 - \overline{x}^2}$

$\sigma_R = \sqrt{\frac{1}{n-1}\sum_{i=1}^n (x_i-\overline{x})^2} = \sqrt{\frac{1}{n-1}\sum_{i=1}^n x_i^2 - \overline{x}^2}$

\end{center}

In order to find the same result in Python as in R, you have to precise ddof = 1 in Python.

:::::: {.columns}
::: {.column width="48%" data-latex="{0.48\textwidth}"}

### Vector manipulation with R

```{r, eval=TRUE}
u <- c(1, 2, 3, 4, 5)
u[2]
u[3:5]
u[5] <- 50
u[1:3] <- 1
u
v <- c(10,20,30,30,60,50)
w <- c(20,10,31,31,61,51)
u <- c(5 ,5 ,5 ,32,62,49)
```

:::
::: {.column width="4%" data-latex="{0.04\textwidth}"}
\ 
<!-- an empty Div (with a white space), serving as
a column separator -->
:::
:::::: {.column width="48%" data-latex="{0.48\textwidth}"}

### Vector manipulation with Python

```{python, eval=TRUE}
u = np.array([1, 2, 3, 4, 5])
u[1]
u[2:5]
u[4] = 50
u[0:3] = 1
u
v = np.array([10,20,30,30,60,50])
w = np.array([20,10,31,31,61,51])
u = np.array([5 ,5 ,5 ,32,62,49])
```

:::
::::::
\newline

#### Equivalence str() R in Python :
The function str() in R show you the structure of your variable, it can be very useful and you'll will find no equivalent function in Python. The only way to get the same information is to create a function :

```{r}
str(u)
```

```{python}
import pandas as pd
du = pd.DataFrame(u)
print(str(du.info()) + "\n"+ str(u))
```

You can see above that the str() function used in the python chunks is not at all the same function as in R. In Python it is used to transform non string type to string character.
\newline
\newline
The one below is for data.frame, keep it for future usage :

```{python, eval = FALSE, error = FALSE}
def rstr(df): 
    structural_info = pd.DataFrame(index=df.columns)
    structural_info['unique_len'] = df.apply(lambda x: len(x.unique())).values
    structural_info['unique_val'] = df.apply(lambda x: [x.unique()]).values
    print(df.shape)
    return structural_info  
rstr(df)
```

\pagebreak

:::::: {.columns}
::: {.column width="48%" data-latex="{0.48\textwidth}"}

### Vector manipulation with R

```{r}
options(width = 30)
sum(is.na(u))
u_ <- c(NA,u,NA,NA)
u_
sum(is.na(u_))
range(u)
range(u_ , na.rm = TRUE)
quantile(u)
summary(u)
sd(u_, na.rm = TRUE)
cor(v,w)
sort(v)
sort(v, decreasing = TRUE)
order(w)
rank(w, ties.method="min")
rank(w, ties.method="max")
pmax(v,w,u)
pmin(v,w,u)
```

:::
::: {.column width="4%" data-latex="{0.04\textwidth}"}
\ 
<!-- an empty Div (with a white space), serving as
a column separator -->
:::
:::::: {.column width="48%" data-latex="{0.48\textwidth}"}

### Vector manipulation with Python

```{python}
np.set_printoptions(
    suppress=True,linewidth=40)
du.isna().sum()
u_ = np.append(u,np.nan)
u_ = np.append(np.nan, u_)
u_ = np.append(u_, np.nan)
u_
du_ = pd.DataFrame(u_)
du_.isna().sum()
print(str(min(u)) + " " + str(max(u)))
print(str(np.nanmin(u_)) + 
  " " + str(np.nanmax(u_)))
np.quantile(u, q = 0.5) # q represent the %
```
```{python, eval = FALSE}
du.describe()   # It works like summary()
```

```{python}
np.nanstd(u_, ddof = 1)
np.corrcoef(v,w)
np.sort(v)
-np.sort(-v)
np.argsort(w)
[sorted(w).index(x) for x in w]
# No easy equivalence for the "max"
```

```{python, include = FALSE}
v = np.array([10,20,30,30,60,50])
w = np.array([20,10,31,31,61,51])
u = np.array([5 ,5 ,5 ,32,62,49])
```

```{python}
z = np.maximum(u,v)
np.maximum(z,w) # Same with minimum()
```


:::
::::::

\newline

:::::: {.columns}
::: {.column width="48%" data-latex="{0.48\textwidth}"}

### Vector manipulation with R

```{r}
options(width = 30)
cumsum(v)
cumprod(v)
cummax(w)
cummin(w)
```

### Boolean operation with R

```{r}
a <- 1 
b <- 2
(a == 1)
(a == b)
(a <= b)
A <- c(TRUE,TRUE,FALSE,FALSE)
B <- c(TRUE,FALSE,TRUE,FALSE)
A & B
A | B
!A
c <- (a > b)
c
v <- c(10,20,30,30,60,50)
t <- (v > 30)
t
w <- v[(v>30)]
w
which(v == 30)
```


:::
::: {.column width="4%" data-latex="{0.04\textwidth}"}
\ 
<!-- an empty Div (with a white space), serving as
a column separator -->
:::
:::::: {.column width="48%" data-latex="{0.48\textwidth}"}

### Vector manipulation with Python

```{python}
np.cumsum(v)
np.cumprod(v)
dw = pd.DataFrame(w)
dw.cummax().transpose()
dw.cummin().transpose()
```

### Boolean operations with Python

```{python, include = FALSE}
a = 1 
b = 2
```


```{python}
bool(a == 1)
bool(a == b)
bool(a <= b)
A = np.array([True,True,False,False])
B = np.array([True,False,True,False])
np.logical_and(A, B)
np.logical_or(A, B)
np.logical_not(A)
c = np.array(a > b)
c
v = np.array([10,20,30,30,60,50])
t = np.array([v > 30])
t
w = v[(v>30)]
w
np.where(v == 30)
```

:::
::::::
\newline

:::::: {.columns}
::: {.column width="48%" data-latex="{0.48\textwidth}"}

### Boolean operations with R

```{r}
which(v == max(v))
which(v == min(v))
s <- 1*t
s
v <- c(10,20,70,30,60,50)
all(v > 5)
any(v < 5)
```

:::
::: {.column width="4%" data-latex="{0.04\textwidth}"}
\ 
<!-- an empty Div (with a white space), serving as
a column separator -->
:::
:::::: {.column width="48%" data-latex="{0.48\textwidth}"}

### Boolean operations with Python

```{python}
np.where(v == max(v))
np.where(v == min(v))
s = 1*t
s
v = np.array([10,20,70,30,60,50])
all(v > 5)
any(v < 5)
```

:::
::::::
\newline

## Conclusion

This conclude the document for now, I'll try to update it if I find useful tips. Hope this will help you in the future.